Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process

### 2024-05-19 report/appendix/fit2data.Rmd
change tutorial to contrained molg

Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data. From each model we obtained 100 simulations (predictions). These can be found in Figure \ref{fig:prediction}. The model predictions fitted the data substantially better for the mixture models. The worse fit was observed for the single distribution Gaussian model which predicted large amounts of negative keystroke intervals which are implausible. Using a log-normal distribution removed these negative keystroke intervals.

```{r eval = F}
theme_set(theme_few(base_size = 9) +
              theme(strip.background = element_blank(),
                    legend.position = "top",
                    legend.justification = "right",
                    panel.grid = element_blank(),
                    panel.background = element_rect(fill = "transparent"), # bg of the panel
                    plot.background = element_rect(fill = "transparent", color = NA),
                    axis.text.y = element_blank(),
                    axis.ticks.y = element_blank(),
                    axis.title = element_blank()))

file <- list.files("../../stanout", pattern = "sims", full.names = T, recursive = T)
ds <- str_remove_all(file, "\\.|\\/|stanout|all_sims.csv") %>% 
    str_to_upper() %>% 
    str_replace(., "PLANTRA", "PLanTra")
ds <- ds[-6]
plots <- list()
plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000, ds = ds[.])) 
plots_fin <- plots[[1]] / plots[[2]] / plots[[3]] / plots[[4]] / plots[[5]] / plots[[6]] + plot_layout(guides = "collect") & theme(legend.position = 'bottom')

plots_fin <- grid.arrange(patchworkGrob(plots_fin), 
                            left = "Density",
                            bottom = "Transition duration in secs")

ggsave("../figures/fitplots.pdf", plots_fin, height = 8.5, width = 6.25)

@@

```{r prediction, fig.cap="Fit to data with models (distribution type) above the plots. Comparison of 50 simulations to observed data summarised by model. For illustration the x-axis was truncated at 2 secs.", eval = T, fig.height=8, fig.width=6.5}
knitr::include_graphics("figures/fitplots.pdf")
```

>>>>

Model fit was assessed by testing to what extent simulations obtained from the model fit the observed data. From each model we obtained 50 simulations (predictions); see Figure \ref{fig:prediction}. The model predictions fitted the data substantially better for the mixture models. The worse fit was observed for the single distribution Gaussian model which predicted a larger number of negative keystroke intervals; a log-normal distribution removed these negative predictions.

```{r eval = F}
theme_set(theme_few(base_size = 10) +
              theme(strip.background = element_blank(),
                    legend.position = "top",
                    legend.justification = "right",
                    panel.grid = element_blank(),
                    panel.background = element_rect(fill = "transparent"), # bg of the panel
                    plot.background = element_rect(fill = "transparent", color = NA),
                    axis.text.y = element_blank(),
                    axis.ticks.y = element_blank(),
                    axis.title = element_blank()))

file <- list.files("../../stanout", pattern = "sims", full.names = T, recursive = T)
ds <- str_remove_all(file, "\\.|\\/|stanout|all_sims.csv") %>% 
    str_to_upper() %>% 
    str_replace(., "PLANTRA", "PLanTra") %>%
    str_replace(., "GUNNEXP2", "GE2")

# remove the shift model
ds <- ds[!str_detect(ds, "SHIFT")]

# get all density plots
plots <- map(1:length(ds), ~get_pdens_plot(file[.], max_iki = 2000))

# remove strips after first row
plots[2:length(ds)] <- map(2:length(ds), ~plots[[.]] + theme(strip.text = element_blank()))

# create labels for datasets
labs <- map(ds, ~ggplot(mapping = aes(x = 0, y = 0, label = .x)) + 
  geom_text(angle = 270, size = 2.5) +
  theme_void())

# add names for data sets on the right
plots_labs <- map(1:length(ds), ~plots[[.]] + labs[[.]] + plot_layout(widths = c(.95, .05)))

# combine legends
plots_ledg <- wrap_plots(plots_labs, guides = "collect", ncol = 1)

# add axe titles
plots_fin <- grid.arrange(patchworkGrob(plots_ledg), 
                            left = "Density", bottom = "Interkey interval in secs")

# save plot 
ggsave("../figures/fitplots.pdf", plots_fin, height = 7.25, width = 6.25)

@@

```{r prediction, fig.cap="Predicted data (50 simulations in each cell) in red in comparison to the interkey interval data shown in black by model types and datasets. Model fit to the data was order from the best (left) to the weakest (right). For illustration the x-axis was truncated at 2 secs.", eval = T, fig.height=8.25, fig.width=6.75}
knitr::include_graphics("figures/fitplots.pdf")
```

### 2024-05-19 report/appendix/l2_constr.Rmd
rearrange markdown sections
change tutorial to contrained molg

## L2 effect (SPL2)

For the SPL2 data (only the `_^[shift]` sentence-transitions) we calculated the L2 effect (i.e. the difference between writing in L2 and L1). The results can be found in Table \ref{tab:l2effect}. The results show longer hesitation duration and more pauses across all transition locations when writing in L2. Only within-word transitions did show negligible evidence for an increased hesitation duration in L2.

```{r l2effect, results = 'asis'}
# cell means and data set difference
cellmeans <- cellmeans %>% 
  select(-`SPL2 (_^[shift] + C)`) %>% 
  pivot_wider(names_from = lang, values_from = `SPL2 (_^[shift])`)

file1 <- list.files(path = "../stanout/spl2", 
                    pattern = "mogbetaco.+.csv$",
                    full.names = T)

@@

write_csv(tmp, "tables/l2_effect_cellmeans_constr.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for language effect. Cell means are shown the slowdown for hesitant transitions when writing in L1 and L2 and the probability of hesitant transitions. The language difference is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "small",
          placement = "b",
          stub_indents = list("Hesitation duration" = 1:3,
                              "Hesitation probability" = 4:6) )

>>>>

# L2 effect (SPL2)

For the SPL2 data we calculated the L2 effect (i.e. the difference between writing in L2 and L1). The results can be found in Table \ref{tab:l2effect}. The results show longer hesitation duration and more pauses across all transition locations when writing in L2. Only within-word transitions did show negligible evidence for an increased hesitation duration in L2.

```{r l2effect, results = 'asis'}
cellmeans <- ps_spl2_both %>% 
  separate(group, into = c("group", "lang"), sep = "; ") %>%
  select(-ds, -group) %>%
  mutate(across(lang, ~str_remove(., "Language: ")),
         across(location, ~str_replace(., "\\n", " "))) %>% 
  pivot_wider(names_from = name, 
              values_from = c(mean, lower, upper)) %>% 
  mutate(across(ends_with("prob"), ~dezero(., 2)),
         across(c(ends_with("beta"),
                  ends_with("delta")), 
                ~as.character(scales::comma(round(.))))) %>% 
  pivot_longer(-c(location, data, lang), 
               names_to = c(".value", "param"),
               names_sep = "_") %>% 
  mutate(est = str_c(mean, " [", lower, ", ", upper,"]")) %>% 
  select(-mean:-upper) %>% 
  pivot_wider(names_from = data, 
              values_from = est) %>% 
  mutate(across(c(location, lang), ~ifelse(param == "beta", "overall", .))) %>% 
  unique()

# cell means and data set difference
cellmeans <- cellmeans %>% 
  select(-`SPL2 (_^[shift] + C)`) %>% 
  pivot_wider(names_from = lang, values_from = `SPL2 (_^[shift])`)

file1 <- list.files(path = "../../stanout/spl2", 
                    pattern = "mogbetaco.+.csv$",
                    full.names = T)

@@

write_csv(tmp, "../tables/l2_effect_cellmeans_constr.csv")

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for language effect. Cell means are shown the slowdown for hesitant interkey intervals when writing in L1 and L2 and the probability of hesitant keyintervals. The language difference is shown on log scale (for durations) and logit scale for probability of hesitant key intervals. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "small",
          placement = "b",
          stub_indents = list("Hesitation showdown" = 1:3,
                              "Hesitation probability" = 4:6) )

### 2024-05-19 report/appendix/masking.Rmd
change tutorial to contrained molg

# Masking effect (CATO, GUNNEXP2)

Studies associated with two datasets (CATO, GUNNEXP2) investigated to what extent masking the previously written text affects keystroke behaviour. The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}. Key transition were longer for masked text than for unmasked text at before and within word transitions but not before sentences. Also dyslexics showed no masking difference at before word transitions. There was no indication of longer or more frequent pauses as a result of masked text.

@@

d_gunn <- read_csv(file) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value)),
         ds = "gunnexp2",
         across(param, ~recode(., "prob" = "theta")),
         group = "non dyslexic") %>% 
  rename(mask = xn)

>>>>

# Masking effect (CATO, GE2)

Studies associated with two datasets (CATO, GE2) investigated to what extent masking the previously written text affects keystroke behaviour. The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}. Key transition were longer for masked text than for unmasked text at before and within word transitions but not before sentences. Also dyslexics showed no masking difference at before word transitions. There was no indication of longer or more frequent pauses as a result of masked text.

@@

d_gunn <- read_csv(file) %>% 
  filter(param %in% c("beta", "delta", "theta")) %>% 
  mutate(across(value, ~ifelse(param == "theta", .*-1, value)),
         ds = "GE2",
         across(param, ~recode(., "prob" = "theta")),
         group = "non dyslexic") %>% 
  rename(mask = xn)

### 2024-05-19 report/appendix/masking_constr.Rmd
change tutorial to contrained molg

# Masking effect (CATO, GUNNEXP2)

Studies associated with two datasets (CATO, GUNNEXP2) investigated to what extent masking the previously written text affects keystroke behaviour. The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}. There was some evidence for longer before-sentence hesitation durations when the text was masked in non-dyslexic writers. Evidence for all other comparisons was negligible.

@@

apa_table(tmp[,-1],
          align = c("l", "l", "l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for masking effect. Cell means are shown for the masked and unmasked writing task in msecs for the slowdown for long transitions and the probability of hesitant transitions. The effect for masking is shown on log scale (for transition durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Hesitation duration" = 1:9,
            "Hesitattion probability" = 10:18) )

>>>>

# Masking effect (CATO, GE2)

Studies associated with two datasets (CATO, GE2) investigated to what extent masking the previously written text affects keystroke behaviour. The mixture model results for the effect of masking is shown in Table \ref{tab:maskingeffect}. There was some evidence for longer before-sentence hesitation durations when the text was masked in non-dyslexic writers. Evidence for all other comparisons was negligible.

@@

apa_table(tmp[,-1],
          align = c("l", "l", "l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for masking effect. Cell means are shown for the masked and unmasked writing task for the slowdown (in msecs) and the probability of hesitant transitions. The effect for masking is shown on log scale (for durations) and logit scale for probability of hesitant transitions. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          longtable = T,
          font_size = "footnotesize",
          stub_indents = list(
            "Hesitation slowdown" = 1:9,
            "Hesitattion probability" = 10:18) )

### 2024-05-19 report/appendix/models.Rmd
change tutorial to contrained molg

We are using the Bayesian framework [e.g. @farrell2018computational; @gelman2014; @lee2014bayesian] to implement 5 statistical models of writing and to evaluate which -- the serial or the cascading view -- captures more successfully the data that arise during keyboard typing. In other words, we use statistical models to map between the data and the theoretically assumed process that generates the data to then compare the predictive power of those models. The models presented in this section build on one another so that later model include assumptions of earlier models.

## Serial model of writing

### Single distribution Gaussian

Under the serial view, all planning must be completed priors to the production onset of the corresponding planning unit. The resulting IKI is sometimes faster or slower depending on, among others, psycholinguistic factors. For example, the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphemes, syllables, and morphemes. There are word-specific factors that influence the IKI that precedes a word but these are beyond the scope of our analysis. We will capture the variability associated with word-features by assuming that before-word IKIs can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average IKI associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the variability associated with factors that we did not further specify in the model. This can be expressed as $\text{iki}_\text{before word} \sim \text{log}\mathcal{N}(\mu, \sigma_\text{e}^2)$. Of importance is the estimated posterior value of $\mu$ as this value captures that time it takes to mentally plan a word.

We can extend this simple model of word-planning to other linguistic location. We introduce earlier that larger linguistic edges are associated with planning on higher levels. For example, at sentence boundaries, planning needs to happen for word-level properties -- which was captured above as average IKI $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07]. Regression formulas can, and commonly are, used to decompose $\mu$ and capture that there is a change in the outcome variable associated with another factors. We can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is the average IKI for word boundaries. However when $\text{x}_\text{sentence}$ takes on the value 1, the average IKI for word boundaries $\alpha$ is incremented by a sentence-related slowdown of $\beta$ msecs. Therefore the value of the $\beta$ parameter represents the additional cognitive demand associated with sentence-initial planning. The application of such a statistical model to the data will then provide us with an estimate of the parameter value that can be used for statistical inference (e.g. whether there is a statistically meaningful difference between IKI associated with words and sentences).

We implemented the average keytransition associated with a given text locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location. Therefore, the model will return one $\beta$ per transition location that capture the posterior distribution of average IKIs. Also the decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is typically called random intercepts $u_\text{participant}$. The random intercepts term $u_\text{participant}$ is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of $\sigma_\text{p}^2$. This value is therefore the difference between the overall posterior estimate and estimate average IKI of a particular participant (i.e. a positive value indicates that a participant is slower than average, a negative value indicates that a participant is faster than average).

The final model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed effects model

@@

where $i$ iterates over every key-transition $i \in 1 \dots N$ with $N$ being the total number of keystroke intervals, and thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.

@@

### Single distribution log-Gaussian

The previous model assumes that the data-generating process is a Gaussian distribution. The next model is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution. There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values. IKIs, as the distance between two subsequent key-down events, must be positive. The lower bound is delimited by a person's ability to move their fingers and keyboard polling. (2) the log-scale is known to be a better match for data from human behaviour and motor responses. In particular, a Gaussian distribution assumes that units are scaled linearly. For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs. This does not necessarily map onto the psychological interpretation for short and long keyintervals. For example effects that result on the motor level within words (e.g. typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2). In other words, although an effect of 25 msecs is large in the context overall fast keyintervals, it is small in the context of overall slow intervals. Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of the IKI scale (motor activity) is more meaningful than a 25 msecs difference on the upper end of the IKI scale (retrieving words, planning sentences).

The model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{log}\mathcal{N}()$ was replaced by $\text{logN}()$.

@@

### Single distribution unequal-variance log-Gaussian

The third model representing the serial view of writing is a single distribution unequal-variance model that, except for the unequal-variance assumption is identical to the model presented in the previous section. The previous models modelled IKIs as a function of transition location so that the estimated average IKI depends on the position of an IKI in the text. The variance associated with the estimated IKIs for each transition location was assumed to be identical (equal variance). This assumption is however not in line with what we know about data from human behaviour. Longer latencies are known to be associated with a larger variances for both response-time data in particular [@wagenmakers2007linear] and human motor behaviour in general [@wing1973response;@schoner2002timing]. For IKIs pauses at larger linguistic edges are plausibly associated with a larger variance because of the larger number of associated processes. Therefore, in equation \ref{eq:unimoduv}, we introduce the assumption that standard deviation $\sigma_{e_\text{location}}^2)$ varies by transition location.

@@

## Parallel cascading model of writing

The following two models are are extensions of the models introduced for the serial view. Crucially, the cascading view allows planning to happen in parallel to production. Therefore, we will reduce the constrain of the serial models that requires all planning to be completed before writing onset. This is done by assuming that IKIs come from a weighted combination of two distributions.

### Two-distributions log-Gaussian mixture model (constrained)

This model extends the assumption of the previous model that processing that involves higher levels of activation lead to longer pauses. Instead of assuming that there is one process that is shifted for IKIs of larger linguistic edges, we introduce the assumption that pauses at larger linguistic edges are more likely but not obligatory. This is achieved by modelling IKIs as coming from a weighted mixture of two distributions associated with two different states:

1. Activation can flow into keystrokes without interruption. These fluent keystroke transitions are merely constrained by a person's ability to move their finger and will be captured by the $\beta$ parameter in equation \ref{eq:bimodcon}. Note that $\beta$ is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon} referring to the same unknown parameter.

2. Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when words or their spelling could not be retrieved in time. The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon}. The magnitude of the slowdown is associated with transition location. This is because delays at larger linguistic units are likely to be associated with higher level planning.  By constraining $\delta$ to be positive, it captures how much longer hesitant IKIs in addition to $\beta$.

\begin{equation}
\begin{aligned}
(\#eq:bimodcon)
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
\text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2\\
		& 0 < \theta < 1
\end{aligned}
\end{equation}

The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant key transitions, and the second line represents fluent key transitions. Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1. $\theta$ is here parameterised to represent the probability that an IKI is associated with the distribution of long IKIs. This probability is inversely related to the mixing weight of the distribution of short IKIs by $1-\theta$ . In other words, a larger weight of one distribution inevitably means a lower weight for the other distribution. The weights of both distributions must sum to 1. We will call this parameter the probability of hesitant transitions.

The probability of hesitant transitions is assumed to vary as a function of both transition location and participants. In line with the literature discussed in the introduction, we assume that pauses are more likely at larger linguistic edges. As pausing is subject to individual differences and writing style (skills), we also assume that some participants pause more at certain transition locations and other participants pause less. This is akin to what is known as a random by-participant slopes model.

Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location. In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing disfluencies is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature]. This was achieved by introducing a parameter $\sigma_\text{diff}$. The consequence is that fluent transitions are assumed to come from a narrower distribution than hesitant transitions.

### Two-distributions log-Gaussian mixture model (unconstrained)

The size of a fluent key-transition does not necessarily vary by transition location. In other words, the parameter $\beta$ is the same for before-sentence, before-word, and within-word transitions. This is what we called a constrained model. However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter (REFERENCE?) or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section). This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions to a space following a word or between a space or shift key press preceding a letter. Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data. We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations.

In this model we assume that $\beta$ varies by transition-location as illustrated in equation \ref{eq:bimoduncon}.

\begin{equation}
\begin{aligned}
(\#eq:bimoduncon)
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
	\text{where: }  & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2\\
		& 0 < \theta < 1
\end{aligned}
\end{equation}

>>>>

We used the Bayesian framework [e.g. @farrell2018computational; @gelman2014; @lee2014bayesian] to implement four statistical models to evaluate which -- the serial or the parallel view -- captures interkey data more successfully. In other words, we use statistical models to map between the keystroke intervals and the theoretically assumed process that generates theses data to then compare the predictive power of those models. These models presented in this section are building on one another so that later models include assumptions of earlier models.

## Single-distribution Gaussian

Under the serial view, all planning must be completed prior to typing onset of the corresponding planning unit. The resulting interkey interval is sometimes faster or slower depending on, among others, psycholinguistic factors; i.e. the interval before a word is shorter for easily retrievable high-frequency words, or longer for low frequency words, shorter for words with fewer graphemes, syllables, and morphemes. For example, we can capture the variability associated with word-features by assuming that before-word interkey intervals can be described as coming from a distribution that is normal (Gaussian) with two parameters, an unknown mean $\mu$, that describes the average interkey interval associated with word-level planning, and a standard deviation $\sigma_\text{e}^2$, that captures the remaining variability associated with factors that we did not further specify in the model. This can be expressed as $\text{iki}_\text{before word} \sim \text{log}\mathcal{N}(\mu, \sigma_\text{e}^2)$ where $\text{iki}_\text{before word}$ are all interkey intervals immediately proceeding a word. Of importance is the estimated posterior value of $\mu$ as this value captures average time needed to mentally prepare a word.

We can extend this simple model of word-planning to other linguistic location. Larger linguistic edges are typically understood as being associated with planning on higher levels. For example, at sentence boundaries, planning needs to happen for word-level properties -- which we captured before as average interkey interval $\mu$ -- but also higher linguistic planning such as clause-level meaning, and dependencies of the sentence-initial noun [@roeser2018advance; @not07]. Using regression formulae we can capture changes in the outcome variable that are associated with other factors. More specifically we can decompose $\mu$ as $\mu = \alpha + \beta \cdot \text{x}_\text{sentence[0,1]}$ so that when $\text{x}_\text{sentence}$ takes on the value 0, the equation reduces to $\mu = \alpha$ which is then the average interkey interval for word boundaries but, when $\text{x}_\text{sentence}$ takes on the value 1, the average interkey interval for word boundaries $\alpha$ is incremented by a slowdown associated with sentence boundaries of $\beta$ msecs. Therefore the value of the $\beta$ parameter represents the additional cognitive demand associated with sentence-initial planning. The application of such a statistical model to the data will therefore provide us with an estimate of the parameter value that can be used for statistical inference (e.g. whether there is a statistically meaningful difference associated with word and sentence boundaries).

We implemented the average interkey interval associated with a given text locations as $\beta_\text{location}$ in equation \ref{eq:unimodgaus} where $\text{location}$ is taking on an index for each transition location. Therefore, the model returns three distributions of $\beta$ values that capture the posterior distribution of average interkey intervals for each transition location. The decomposition of $\mu$ allows us to address the fact that some writers are faster than other writers by introducing a parameter for what is called random intercepts $u_\text{participant}$. The random intercepts term $u_\text{participant}$ is constrained so that the value it takes on comes from a normal distribution with a mean of 0 and a standard deviation of $\sigma_\text{p}^2$. The $u_\text{participant}$ value is therefore the difference between the overall posterior estimate and estimate average interkey interval of a particular participant (i.e. a positive value indicates that a participant is slower than average; a negative value indicates that a participant is faster than average).

The resulting model is shown in equation \ref{eq:unimodgaus} and represents a Gaussian mixed-effects model

@@

where $i$ iterates over every interkey interval $i \in 1 \dots N$ with $N$ being the total number of interkey intervals; thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.

@@

## Single distribution log-Gaussian

The previous model assumes that the data-generating process follows a Gaussian distribution. As a consequence the Gaussian model generates values that are symmetrically distributed and, in principle, allows values to be negative. The model presented in this section is largely identical to the previous model but instead of assuming a Gaussian, we assume that the data come from a log-normal (log-Gaussian) distribution.

There are, at least, two arguments for using a log-Gaussian distribution: (1) log-Gaussians are zero-bound; in contrast to Gaussians, a log-Gaussian does not allow negative values. This is a desirable property because the distance between two subsequent key events must be positive. Instead the lower bound is delimited by a person's ability to move their fingers and keyboard polling. (2) the log-scale is known to be a better match for data from human behaviour and motor responses. In particular, a Gaussian distribution assumes that units are scaled linearly. For example, a difference of 25 msecs is the same between 100 and 125 msecs as between 5 secs and 5,025 msecs. This does not map onto the psychological interpretation for short and long interkey intervals. For example effects that result from the motor level within words (e.g. typing a high vs a low frequency bigram) are smaller than differences that are due to high levels of processing (retrieving a word in an L1 or L2). In other words, an effect of 25 msecs can be large in the context of fast inerkey intervals but it is small in the context of overall slow intervals. Log-Gaussian distributions are a natural way of translating a linear scale to an exponential scale so that a 25 msecs difference on the lower end of interkey intervals (motor activity) is more psychologically meaningful than a 25 msecs difference on the upper end (retrieving words, planning sentences).

This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{log}\mathcal{N}()$ was replaced by $\text{log}\mathcal{N}()$.

@@

## Single distribution unequal-variance log-Gaussian

The third model is identical to the model presented in the previous section but removed the equal-variance assumption for the width of the distributions assocated with transition locations. The previous models modelled interkey intervals as a function of transition location meaning that the average interkey interval depends on its position in the text. The variance associated with the estimated interkey intervals for each transition location was assumed to be identical (equal variance). This assumption is however not in line with what we know about data from human motor behaviour. Longer latencies are known to be associated with a larger variances for both response-time data in particular [@wagenmakers2007linear] and human motor behaviour in general [@wing1973response;@schoner2002timing]. For interkey intervals pauses at larger linguistic edges can therefore be assumed to be associated with a larger variance also because of the larger number of associated processes. Therefore, in equation \ref{eq:unimoduv}, we introduced the assumption that standard deviations $\sigma_{e_\text{location}}^2)$ vary by transition location.

@@

## Two-distributions log-Gaussian mixture model

The following model is an extension of the last single-distribution model. In contrast to the models introduced for the serial view, the cascading view allows planning to unfold in parallel to production. Therefore, we removed for the model presented in this section the constraint that all planning must be completed before writing onset. Instead of assuming that larger linguistic edges shift the distribution of interkey intervals as in the previous models, we modelled pauses in the distribution of interkey intervals, under the parallel view, to be more likely to occur at larger linguistic edges. This was achieved by introducing the model assumption that interkey intervals come from a weighted combination (mixure) of two distributions associated with two different states:

1. Activation can flow into keystrokes without interruption. These fluent keystroke transitions are merely constrained by a person's ability to move their finger and were captured by $\beta$ in equation \ref{eq:bimodcon}. Note that $\beta$ is represented in both log-Gaussian distributions in equation \ref{eq:bimodcon} referring to the same unknown parameter.

2. Interruptions in the activation flow from higher to lower levels result in longer keystroke intervals, when information, for example words or their spelling, could not be retrieved in time. The slowdown for these hesitant transitions is captured by $\delta$ in the first line of equation \ref{eq:bimodcon} which represents the magnitude of the slowdown associated with transition location. The slowdown $\delta$ was allowed to vary by transition locations because hesitations at larger linguistic units are likely to be associated with higher level planning which causes longer delays in the output. The $\delta$ parameter was constrained to be positive, so that it captures how much longer hesitant interkey intervals are in addition to $\beta$.

\begin{equation}
\begin{aligned}
(\#eq:bimodcon)
\text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i], participant[i]}) \times \text{log}\mathcal{N}(\beta + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\
\text{where: } & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2\\
		& 0 < \theta < 1
\end{aligned}
\end{equation}

The first line of equation \ref{eq:bimodcon} represents the distribution of hesitant interkey intervals; the second line represents fluent interkey intervals. Each of these two distributions is associated with the mixing weight $\theta$ which is a proportion that is constrained to be larger than 0 and smaller than 1. $\theta$ is here parameterised to represent the probability that an interkey interval is associated with the distribution of long interkey intervals, hence representing the hesitation probability. This probability is inversely related to the mixing weight of the distribution of short interkey intervals by $1-\theta$ because the weights of both distributions must sum to 1. In other words, a larger weight for one distribution inevitably means a lower weight for the other distribution.

The probability of hesitant transitions is assumed to vary as a function of both transition location and participants. In line with the literature discussed in the introduction, we assumed that pauses are likely to vary across linguistic edges. As pausing is subject to individual differences and writing style (skills), we also assumed that some participants pause more at certain transition locations and other participants pause less.

<!-- This is akin to what is known as a random by-participant slopes model.  -->

Lastly, we carried over the unequal variance assumption and let the standard deviations $\sigma_{e'}^2$ and $\sigma_{e}^2$ vary by transition location. In addition we constrained the variances so that $\sigma_{e'}$ associated with the distribution of typing hesitations is larger than the variance associated with fluent transitions $\sigma_e$ [see @vasishth2017; @vasishth2017feature].

<!-- ### Two-distributions log-Gaussian mixture model (unconstrained) -->

<!-- The size of a fluent key-transition does not necessarily vary by transition location. In other words, the parameter $\beta$ is the same for before-sentence, before-word, and within-word transitions. This is what we called a constrained model. However, letter bigrams and trigrams may be systematically executed faster than transitions between between space and a letter or complex keystrokes that comprise space and shift-letter combinations for upper case characters before sentences (we will address the latter possibility in the results section). This is because bigrams / trigrams might be stored, retrieved and executed as motor codes but not transitions to a space following a word or between a space or shift key press preceding a letter. Also, because of the necessarily larger number of within-word transitions, as opposed to before-word and sentence transitions, the posterior of the constrained model is dominated by within-word transition data. We therefore also implemented an unconstrained model that assumes that the size of fluent transitions varies across transition locations. -->

<!-- In this model we assume that $\beta$ varies by transition-location as illustrated in equation \ref{eq:bimoduncon}. -->

<!-- \begin{equation} -->
<!-- \begin{aligned} -->
<!-- (\#eq:bimoduncon) -->
<!-- \text{iki}_{i} \sim\text{ } & \theta_\text{location[i], participant[i]} \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + \delta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e'_\text{location[i]}}^2) + \\ -->
<!--   & (1 - \theta_\text{location[i], participant[i]}) \cdot \text{log}\mathcal{N}(\beta_\text{location[i]} + u_\text{participant[i]}, \sigma_{e_\text{location[i]}}^2)\\ -->
<!-- 	\text{where: }  & u_\text{participant[i]} \sim \mathcal{N}(0, \sigma_\text{p}^2) \\ -->
<!-- \text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\ -->
<!-- 		& \sigma_{e'}^2 > \sigma_{e}^2\\ -->
<!-- 		& 0 < \theta < 1 -->
<!-- \end{aligned} -->
<!-- \end{equation} -->
